Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-22 VegOzone.tex
Edited conclusions

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}.

@@

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, this indicates a possibly considerable underestimation of ozone risk for subarctic species. Especially, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the ozone damage risk for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later onset and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated for a specific acclimation and is of the same order of magnitude as the divergence between the hypothetical acclimations. For comparison, we averaged the corresponding biomass reduction for the MM temperature and light response functions with both SWP on/off but with bespoke GS over both years, we find a non negligible, which means above systematic uncertainties, low-bias for both downy birch and perennial grassland with respect to the MM default parameterization. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

@@

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found onsets ($106/122\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the onset of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods to estimators show a later onset for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

@@

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrics like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periods of time. Consequently, acute cellular damage caused by very high fluxes over a short period of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differentiate picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, were found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandic vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need verification by experimental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite might be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch and perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier onset of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start of the growing season. It is therefor not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier onset of the growing season, will be reduced compared to present day. Hence, a simulated earlier onset of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions have been evaluated based on central European species or PFTs and might as well be subject to biases.

@@

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature response function may have important indications for land-surface modeling, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out with respect to default perennial grassland parameterization that that did not show any substantial stomatal conductance in 2019 which we characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

>>>>

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated start of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}.

@@

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed start of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, this indicates a possibly considerable underestimation of ozone risk for subarctic species. Especially, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the ozone damage risk for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold types with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated for a specific acclimation and is of the same order of magnitude as the divergence between the hypothetical acclimations. For comparison, we averaged the corresponding biomass reduction for the MM temperature and light response functions with both SWP on/off but with bespoke GS over both years and find a non negligible, which means above systematic uncertainties, low-bias for both downy birch and perennial grassland with respect to the MM default parameterization. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

@@

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found start of growing season ($106/122\,\unit{doy}$) is within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

@@

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were, in general, not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrics like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periods of time. Consequently, acute cellular damage caused by very high fluxes over a short period of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differentiate picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, was found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandic vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation ought to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need verification by experimental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed start of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite is be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch and perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier start of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start of the growing season. It is therefor not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier start of the growing season, will be reduced compared to present day. Hence, a simulated earlier start of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions were established based on central European species or PFTs and might as well be subject to biases.

@@

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.\\

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature response functions, once varified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.
